Determinism, Complexity, and Predictability in Computer Performance

### 2014-07-24 concl.tex
airplane edits

@@

In this paper, we have argued that redundancy~\cite{crutchfield2003}
is a useful definition of the inherent predictability of an empirical
time series.  To operationalize that definition, we use an
approximation of the Kolmogorov-Sinai entropy~\cite{lind95}, estimated
using a weighted version of the permutation entropy
of~\cite{bandt2002per}.  This WPE technique---an ordinal calculation
of forward information transfer in a time series---is ideal for our
purposes because it works with real-valued data and is known to
converge to the true entropy value. Using a variety of forecast models
and more than a hundred time-series data sets from a
computer-performance experiment, we have shown that prediction
accuracy is indeed correlated with weighted permutation entropy: the
higher the WPE, in general, the higher the prediction error.  The
relationship is roughly logarithmic.  [[This is just an observed result.
It makes sense because, firstly, any signal with zero entropy should
be perfectly predictable, and so should ideally have a MASE value of
zero.  As the entropy rises, the MASE value rise: slowly at first
because the signal still has lots of info, then more rapidly because
[[why]], and finally saturating at [[what]] because of [[why]].
Supporting our hypothesis, forecast errors for the \col and \svdfive
signals are higher than the corresponding WPEs suggest---except for
the nonlinear LMA predictor.  There are some exceptions: \svdone is
easier to predict than its WPE value suggests using any of the three
forecast strategies studied here.  On the whole, though, signals with
higher weighted permutation entropy are harder to predict.

@@

An important practical corollary to this empirical correlation of
predictability and WPE is a practical strategy for assessing
appropriateness of forecast methods.  If the forecast produced by a
particular method is poor but the time series contains a significant
amount of predictive structure, one can reasonably conclude that that
method is inadequate to the task and that one should seek another
method.  The nonlinear LMA method, for instance, performs better in
most cases because it is more general.  (This is particularly apparent
in the \col and \svdfive examples.)
% , where the other two methods do not perform well.)
The \naive ~method, which simply predicts the mean, can work very well
on noisy signals because it effects a filtering operation.  The simple
random-walk strategy outperforms LMA, \arima, and the \naive
~method on the \gcc signal, which is extremely complex---i.e.,
extremely low redundancy.
%  \naive ~wins on \svdone with the exception to the 5 outliers.
The dashed curve and shaded region in
Figures~\ref{fig:wpe_vs_mase_best} and~\ref{fig:wpe_vs_mase_all}
operationalize the discussion in this paragraph.  These geometric
features are a preliminary, but potentially useful, empirical
guideline for knowing when a model is not well-matched to the task at
hand: [[a MASE score that is significantly higher than $3.58 \cdot
    \mathrm{WPE} - 1.91$, where $\mathrm{WPE}$ is that of the
    corresponding signal]] [[need to reword]], is an indication that
that time series has more predictive structure than that forecast
model can capture and exploit.

>>>>

[[Moved this out of intro but it was redundant in the model section
    and some of it is repeated in the results section.  Maybe here?
    The forecast methods used in this study were chosen to be a
    representative sampling of standard prediction strategies, but
    they do not, of course, cover that space exhaustively.  Our goal
    here is an empirical assessment of the relationship between
    predictability and complexity, not formal results about a ``best''
    predictor for a given time series.  From a practitioner's
    standpoint, it would be useful to know \emph{ a priori} if a time
    series contains enough predictive structure to make it worth
    spending the time and effort searching for a good forecast method.
    It would also be useful to know if a given method is
    inadequate---that is, if an alternative method could do better and
    so one should continue searching.]]

@@

In this paper, we have argued that redundancy~\cite{crutchfield2003}
is a useful definition of the inherent predictability of an empirical
time series.  To operationalize that definition, we use an
approximation of the Kolmogorov-Sinai entropy~\cite{lind95}, estimated
using a weighted version of the permutation entropy
of~\cite{bandt2002per}.  This WPE technique---an ordinal calculation
of forward information transfer in a time series---is ideal for our
purposes because it works with real-valued data and is known to
converge to the true entropy value. Using a variety of forecast models
and more than a hundred time-series data sets from a
computer-performance experiment, we have shown that prediction
accuracy is indeed correlated with weighted permutation entropy: the
higher the WPE, in general, the higher the prediction error.  The
relationship is roughly logarithmic.

[[Maybe put (or reiterate) the ``range of examples'' material here.
    This is just an observed result.  It makes sense because any
    signal with zero entropy should be perfectly predictable, and so
    should ideally have a MASE value of zero.  As the entropy rises,
    the MASE value rise: slowly at first---where the signal is not
    very complex---and then more rapidly after WPE $\approx$ 0.5.  We
    are in the process of exploring the theory behind this shape, as
    well as testing other shapes.  Supporting our hypothesis, forecast
    errors for the \col and \svdfive signals are higher than the
    corresponding WPEs suggest---except for the nonlinear LMA
    predictor.  There are some exceptions: \svdone is easier to
    predict than its WPE value suggests using any of the three
    forecast strategies studied here.  On the whole, though, signals
    with higher weighted permutation entropy are harder to predict.

@@

An important practical corollary to this empirical correlation of
predictability and WPE is a practical strategy for assessing
appropriateness of forecast methods.  If the forecast produced by a
particular method is poor but the time series contains a significant
amount of predictive structure, one can reasonably conclude that that
method is inadequate to the task and that one should seek another
method.  The nonlinear LMA method, for instance, performs better in
most cases because it is more general.  (This is particularly apparent
in the \col and \svdfive examples.)
% , where the other two methods do not perform well.)
The \naive ~method, which simply predicts the mean, can work very well
on noisy signals because it effects a filtering operation.  The simple
random-walk strategy outperforms LMA, \arima, and the \naive
~method on the \gcc signal, which is extremely complex---i.e.,
extremely low redundancy.
%  \naive ~wins on \svdone with the exception to the 5 outliers.
The dashed curve and shaded region in
Figures~\ref{fig:wpe_vs_mase_best} and~\ref{fig:wpe_vs_mase_all}
operationalize the discussion in this paragraph.  These geometric
features are a preliminary, but potentially useful, empirical
guideline for knowing when a model is not well-matched to the task at
hand: a MASE score that is significantly\footnote{more than one
  $\sigma$ out} higher than [[$3.58 \cdot \mathrm{WPE} - 1.91$]],
where $\mathrm{WPE}$ is that of the corresponding signal, is an
indication that that time series has more predictive structure than
that forecast model can capture and exploit.

### 2014-07-24 intro.tex
airplane edits

This leads naturally to an important and challenging question: given a
noisy real-valued empirical time series, does there exist a forecast
model that can leverage the information (if any) that is being
transmitted from past to future by the underlying generating process?
A first step in answering this question is to reliably quantify where
on the complexity spectrum a given time series falls; a second step is
to determine how complexity and predictability are related in these
kinds of data sets.  With these answers in hand, one can develop a
practical strategy for assessing appropriateness of forecast methods
for a given time series.  If the forecast produced by a particular
method is poor, for instance, but the time series contains a
significant amount of predictive structure, one can reasonably
conclude that that method is inadequate to the task and that one
should seek another method.  The goal of this paper is to develop
effective, practical heuristics for that strategy.

@@

\item The permutation entropy of a noisy real-valued time series is
  correlated with the accuracy of an appropriate predictor.

\item The relationship between permutation entropy and prediction
  accuracy is a useful empirical heuristic for identifying when there
  is a mismatch between a time series from an unknown system and a
  given prediction model.
\end{enumerate}
% 
There has, of course, been a great deal of good work on different ways
to measure the complexity of data, and previous explorations have
confirmed repeatedly that complexity is a challenge to prediction.  It
is well known that the way information is generated and processed
internally by a system plays a crucial role in the success of
different forecasting methods---and in the choice of which method is
appropriate for a given time series.  This constellation of issues has
not been properly explored, however, in the context of noisy, poorly
sampled, real-world data from unknown systems.  That exploration---and
the development of strategies for putting its results into effective
practice---is the primary contribution of this work.  The empirical
results in the later sections of this paper not only elucidate the
relationship between complexity and predictability.  The strategy that
we derive from these results can also aid practitioners in choosing an
appropriate prediction model for a given real-world noisy time series
from an unknown system---a challenging task for which little guidance
is currently available.

>>>>

This leads naturally to an important and challenging question: given a
noisy real-valued empirical time series, does there exist a forecast
model that can leverage the information (if any) that is being
transmitted from past to future by the underlying generating process?
A first step in answering this question is to reliably quantify where
on the complexity spectrum a given time series falls; a second step is
to determine how complexity and predictability are related in these
kinds of data sets.  With these answers in hand, one can develop a
practical strategy for assessing appropriateness of forecast methods
for a given time series.  If the forecast produced by a particular
method is poor, for instance, but the time series contains a
significant amount of predictive structure, one can reasonably
conclude that that method is inadequate to the task and that one
should seek another method.  The goal of this paper is to develop
effective heuristics to put that strategy into practice.

@@

\item The permutation entropy of a noisy real-valued time series from
  an unknown system is correlated with the accuracy of an appropriate
  predictor.

\item The relationship between permutation entropy and prediction
  accuracy is a useful empirical heuristic for identifying mismatches
  between prediction models and time-series data.

\end{enumerate}
% 
There has, of course, been a great deal of good work on different ways
to measure the complexity of data, and previous explorations have
confirmed repeatedly that complexity is a challenge to prediction.  It
is well known that the way information is generated and processed
internally by a system plays a crucial role in the success of
different forecasting methods---and in the choice of which method is
appropriate for a given time series.  This constellation of issues has
not been properly explored, however, in the context of noisy, poorly
sampled, real-world data from unknown systems.  That exploration---and
the development of strategies for putting its results into effective
practice---is the primary contribution of this work.  The empirical
results in the later sections of this paper not only elucidate the
relationship between complexity and predictability.  The strategy that
we derive from these results can also aid practitioners in choosing an
appropriate prediction model for a given real-world noisy time series
from an unknown system---a challenging task for which little guidance
is currently available.

### 2014-07-24 methods.tex
airplane edits

The time-series data sets for these experiments were collected on an
Intel Core\textsuperscript{\textregistered} i7-2600-based machine.
% 
% running the 2.6.38-8 Linux kernel.  This particular microprocessor
% chip has eight processing units, a clock rate of 3.40 GHz, and a cache
% size of 8192 KB.  
% 
We gathered performance traces during the execution of three different
programs---the simple \col loop whose performance is depicted in
Figure~\ref{fig:col-ipc}, as well as two more-complex programs: one
from the SPEC 2006CPU benchmark suite (\gcc), and one from the LAPACK
linear algebra package (\svd).  In these experiments, the scalar
observation $x_i$ was a measurement of the processor performance at
time $i$ during the execution of each program.  
% 
% To record these measurements, we used the {\tt libpfm4} library, via
% PAPI
% % 
% % \footnote{Performance Application Programming Interface}
% % 
% 5.2~\cite{papi}, to stop program execution at 100,000-instruction
% intervals---the unit of time in this paper---and read the contents of
% the CPU's onboard hardware performance monitors\footnote{These
%   specialty registers are built into modern microprocessors for the
%   purpose of monitoring and storage of performance data.  They can be
%   programmed to measure a wide array of processor and memory metrics
%   besides the one used here.}, which we had programmed to count how
% many instructions were executed in each clock cycle (IPC).  
% 
For an in-depth description of this experimental setup,
% including discussion of the implications of the sampling interval,
please see \cite{zach-IDA10,mytkowicz09,todd-phd}.  For statistical
validation, we collected 15 performance traces from each of the three
programs.  These traces, and the processes that generated them, are
described in more depth in the rest of this section.

@@

\svd is a Fortran program from the LAPACK linear algebra package
\cite{lapack}.  It calculates the singular value decomposition of a
rectangular $M$ by $N$ matrix with real-valued entries.  For our experiments, we
choose $M=750$ and $N=1000$ and generate the matrix entries randomly.
% 
% \footnote{Multiple randomly generated matrices were
%   investigated but no measurable effect was present in the resulting
%   time series.}. 
% 
The behavior of this program as it computes the singular values of
this matrix is very interesting, as shown in Figure
\ref{fig:svd-ts-colored}.  As the code moves though its different
phases---diagonalizing the matrix, computing its transpose,
multiplying, etc.---the processor utilization patterns change quite radically,
as shown in Figure~\ref{fig:svd-ts-colored}.
\begin{figure}[t]
    \centering
    \includegraphics[width=\columnwidth]{figs/SVD1RegimesColored}
    \caption{An IPC trace during the execution of \gcc.  The colors
      identify the different \emph{segments} of the signal that are
      discussed in the text.}
    \label{fig:svd-ts-colored}
  \end{figure}
For the first $\sim$21,000 measurements (21,000 $\times$ 100,000
instructions), roughly 1.8 instructions are executed per cycle, on the
average, by the eight processing units on this chip.  After that, the
IPC moves through a number of different oscillatory regimes, which we
have color-coded in the figure in order to make textual
cross-references easy to track.

This wide range of behaviors provides a unique advantage, for the
purposes of this paper, in that a number of different generating
processes---with a wide range of complexities---are at work in
different phases of a single time series.  The complexity of \col and
\gcc in Figures~\ref{fig:col-ipc} and \ref{fig:gcc-ts} appears to be
far more consistent over time---probably the result of a single
generating process.  \svd, in contrast, has multiple regimes, each the
result of a different generating process.  To take advantage of this,
we split the signal into six different segments, thereby obtaining an
array of examples for the analyses in the following sections.  For
notational convenience, we refer to these 90 time-series data
sets\footnote{15 runs, each with six regimes} as {\tt dgesdd$_i$},
with $i \in \{1\dots6\}$ where $i$ corresponds to one of the six
segments of the signal, ordered from left to right.  These segments,
which were determined visually, are shown in different colors in
Figure~\ref{fig:svd-ts-colored}.  Visual decomposition is subjective,
of course, particularly since the regimes exhibit some fractal
structure.  Thus, it may be the case that more than one generating
process is at work in each of our segments.  This is a factor in the
discussion of Section~\ref{sec:results}.

>>>>

The time-series data sets for these experiments were collected on an
Intel Core\textsuperscript{\textregistered} i7-2600-based machine.
% 
% running the 2.6.38-8 Linux kernel.  This particular microprocessor
% chip has eight processing units, a clock rate of 3.40 GHz, and a cache
% size of 8192 KB.  
% 
We gathered performance traces during the execution of three different
programs---the simple \col loop whose performance is depicted in
Figure~\ref{fig:col-ipc}, as well as two more-complex programs: one
from the SPEC 2006CPU benchmark suite (\gcc), and one from the LAPACK
linear algebra package (\svd).  In all of these experiments, the
scalar observation $x_i$ was a measurement of the processor
performance at time $i$ during the execution of each program.
% 
% To record these measurements, we used the {\tt libpfm4} library, via
% PAPI
% % 
% % \footnote{Performance Application Programming Interface}
% % 
% 5.2~\cite{papi}, to stop program execution at 100,000-instruction
% intervals---the unit of time in this paper---and read the contents of
% the CPU's onboard hardware performance monitors\footnote{These
%   specialty registers are built into modern microprocessors for the
%   purpose of monitoring and storage of performance data.  They can be
%   programmed to measure a wide array of processor and memory metrics
%   besides the one used here.}, which we had programmed to count how
% many instructions were executed in each clock cycle (IPC).  
% 
% These traces, and the processes that generated them, are described
% briefly in the rest of this section.
% 
For statistical validation, we collected 15 performance traces from
each of the three programs.  For an in-depth description of this
experimental setup,
% including discussion of the implications of the sampling interval,
please see \cite{zach-IDA10,mytkowicz09,todd-phd}.

@@

\svd is a Fortran program from the LAPACK linear algebra package
\cite{lapack}.  It calculates the singular value decomposition of a
rectangular $M$ by $N$ matrix with real-valued entries.  For our experiments, we
choose $M=750$ and $N=1000$ and generate the matrix entries randomly.
% 
% \footnote{Multiple randomly generated matrices were
%   investigated but no measurable effect was present in the resulting
%   time series.}. 
% 
The behavior of this program as it computes the singular values of
this matrix is very interesting, as is clearly visible in Figure
\ref{fig:svd-ts-colored}.  As the code moves though its different
phases---diagonalizing the matrix, computing its transpose,
multiplying, etc.---the processor utilization patterns change quite radically,
as shown in Figure~\ref{fig:svd-ts-colored}.
\begin{figure}[t]
    \centering
    \includegraphics[width=\columnwidth]{figs/SVD1RegimesColored}
    \caption{An IPC trace during the execution of \gcc.  The colors
      identify the different \emph{segments} of the signal that are
      discussed in the text.}
    \label{fig:svd-ts-colored}
  \end{figure}
For the first $\sim$21,000 measurements (21,000 $\times$ 100,000
instructions), roughly 1.8 instructions are executed per cycle, on the
average, by the eight processing units on this chip.  After that, the
IPC moves through a number of different oscillatory regimes, which we
have color-coded in the figure in order to make textual
cross-references easy to track.

The wide range of behaviors in Figure~\ref{fig:svd-ts-colored}
provides a unique advantage, for the purposes of this paper, in that a
number of different generating processes---with a wide range of
complexities---are at work in different phases of a single time
series.  The complexity of \col and \gcc in Figures~\ref{fig:col-ipc}
and \ref{fig:gcc-ts} appears to be far more consistent over
time---probably the result of a single generating process.  \svd, in
contrast, has multiple regimes, each the result of a different
generating process.  To take advantage of this, we split the signal
into six different segments, thereby obtaining an array of examples
for the analyses in the following sections.  For notational
convenience, we refer to these 90 time-series data sets\footnote{15
  runs, each with six regimes} as {\tt dgesdd$_i$}, with $i \in
\{1\dots6\}$ where $i$ corresponds to one of the six segments of the
signal, ordered from left to right.  These segments, which were
determined visually, are shown in different colors in
Figure~\ref{fig:svd-ts-colored}.  Visual decomposition is subjective,
of course, particularly since the regimes exhibit some fractal
structure.  Thus, it may be the case that more than one generating
process is at work in each of our segments.  This is a factor in the
discussion of Section~\ref{sec:results}.
